Discussion Draft

### 2020-04-01 doc/Discussion.rmd
more thoughts

Under the spindle based model faster male specific evolution of the genome wide recombination rate is primarily driven by differences in the strength of the spindle assemby checkpoint (SAC). (we won't go deeply into the molecular details (but the genes and pathways are will known, we refer the reader to [@laneKauppi2019]).

This is our logic for how a effective SAC will caus puriying selection on the bivalent / chromosome structure at meiotisis I division / reuctional division (MI).

The transition from metaphse to analphase is irreversable, so everything (all bivalents in the cell) has to be set up correctly before anaphase to prevent lower fitness later in the gametogenesis pathway. If unstable bivalents are detected on the MI spindle, the SAC delays the transition and buys time for the lose bivalent to correctly attach to the MT spindle. If this is delayed long enough, apoptosis will be triggered. If there are genetic variants which increase the number of unstable bivalents, they selected against in favor to genetic variants causing faster stabilization of bivalents on the MI spindle. Thus the SAC can lead to stabilizing selection on the chromosome structure.

@@

While the modifier models parametrize sexual dimorphism in the strength of selection which can shape the variance in a trait these types of models make predictions  for variance between individuals but not within individuals where the genetics are constant. (selection on penetrance).
In the spindle based selection model,  the source of between cell variance is attributed to difference strength of checkpoints. (Sexual dimorphism in the stringency of the spindle assembly checkpoint is a well established result in the literature. Most empirical results come from mammals (is this true?) yet, some of the hypotheses for the mechanism of inefficient SAC in oocytes are attributed to acentrosomal spindles (cite) and cell volume (cite), which are conserved sexual dirmohpic features of gametogenesis (cite).

>>>>

Under the spindle based model faster male specific evolution of the genome wide recombination rate is primarily driven by differences in the strength of the spindle assemby checkpoint (SAC). (we won't go deeply into the molecular details (but the genes and pathways are will known, we refer the reader to [@laneKauppi2019]). The transition from metaphse to analphase is irreversable, so all bivalents in the cell has to be set up correctly before anaphase to prevent lower fitness later in the gametogenesis pathway. If unstable bivalents are detected on the MI spindle, the SAC delays the transition and buys time for the lose bivalent to correctly attach to the MT spindle. If this is delayed long enough, apoptosis will be triggered. If there are genetic variants which increase the number of unstable bivalents, they selected against in favor to genetic variants causing faster stabilization of bivalents on the MI spindle. Thus the SAC can lead to stabilizing selection on the chromosome structure.

@@

While the modifier models parametrize sexual dimorphism in the strength of selection which can shape the variance in a trait these types of models make predictions  for variance between individuals but not within individuals where the genetics are constant. (selection on penetrance).

In the spindle based selection model, the source of between cell variance is attributed to difference strength of checkpoints.
As mentioned above, there is sexual dimorphism in the stringency of the spindle assembly checkpoint (SAC).  And these are linked to conserved features of gametogenesis.

As mentioned above (fundamental/conserved) sex differences in the MI spindle presence of centrioles in males anchors spindle at two cell poles. While oocytes lack centrioles and so the MT spindle is more difuse and made of mulitple microtubule organizing centers (aMTOCs) (Schuh Ellenburg 2007). This or larger cell volumne might -- lead to anaphase promoting signal being more difuse (or tension signal?)

2. Greater between cell variation in females, leaky SAC in oocytes can cause more between cell variation within an individual.

2a. relaxed selection on chromosome structure in oocyte / leaky SAC, can lead to larger variance in crossover number between cells. The opposite dynamics of the purifying selection for 'strict' SAC in spermatocytes.

2b. (leaky SAC could be adaptive/prevent to selfish genetic elements (Lampson))?
